Transit Delays in the Toronto Transit Commission

### 2024-01-26 outputs/paper/paper.qmd
Finished all the paper writing.

---
title: "Exploring TTC Delays"
subtitle: "A Multimodal Analysis"
author: Luca Carnegie
thanks: "Code and data are available at: https://github.com/lcarnegie/OpenDataToronto. Thank you to Nescafé coffee, Rajan Maghera, Hannah Yu, and Sehar Bajwa for your love and support - could not have done it without you. No thanks go to Rohan in particular for not pointing out all the R he wanted us to learn that was in his textbook... sigh. I will try the appendix cocktails someday, though. :)"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
toc: true
fig_caption: yes
number-sections: true
bibliography: references.bib

@@

# Data {#sec-data}
```{r data import}
#| message: false
#| echo: false
#| label: import data
#| tbl-cap: Sample of Cleaned Delay Data
#| fig-align: center

@@

## Incident Type {#subsec1-data}
```{r ggplot Common Delay Reasons}
#| message: false
#| echo: false
#| label: common-reasons-delay
#| tbl-cap: Most Common Reasons for Delay on TTC

@@

```
Subways didn't have any reason for delay, yet have the most incidence of delay.

## Line/Route {#subsec2-data}
```{r ggplot top 10 routes with longest avg. delay}
#| message: false
#| echo: false
#| label: lines-longest-delays
#| tbl-cap: Incidence of Delay, by Route/Line

```{r ggplot top 10 routes with most delay incidence}
#| message: false
#| echo: false
#| label: lines-most-delays
#| tbl-cap: Most Common Reasons for Delay on TTC

avg_time_delayed_vehicle |>
  ggplot(mapping = aes(x = vehicle, y = avg_delay)) +
  geom_col() + 
  labs(title="Average Delay by Vehicle", x="Vehicle Type", y="Average Delay (mins)") +
  theme_minimal()
```
```{r ggplot transport with most incidence of delays }
#| message: false
#| echo: false
#| label: most-incidents-vehicle

num_incidents_vehicle |>
  ggplot(mapping = aes(x = vehicle, y = num_incidents)) +
  geom_col() + 
  labs(title="Number of Incidents by Vehicle", x="Vehicle Type", y="Number of Incidents") +
  theme_minimal()

## Day of Week {#subsec4-data}
```{r ggplot Day of Week with Delay incidence}
#| message: false
#| echo: false
#| label: transport-delay-time
#| tbl-cap: Most Common Reasons for Delay on TTC
#Day of week with most incidence of delay

@@

num_incidents_vehicle |>
  ggplot(mapping = aes(x = Day, y = num_incidents)) +
  geom_col() +
  labs(title="Incidence of Delay by Day of Week", x="Day of Week", y="Incident Count") +
  theme_minimal()

@@

@@

#plot
avg_time_delayed_vehicle |>
  ggplot(mapping = aes(x = Day, y = avg_delay)) +
  geom_col() + 
  labs(title="Average Delay Time by Day of Week", x="Vehicle Type", y="Average Delay (mins)") +
  theme_minimal()
```

## Time of Year {#subsec5-data}
```{r ggplot avg delay change monthly}
#| message: false
#| echo: false
#| warning: false
#| label: monthly-delay-times
#| tbl-cap: Average Monthy Delay Times
#Line chart of how avg. delay time changes over the months

@@

# plot the graph
avg_delay |>
  ggplot(aes(x=month, y=`Min Delay`, color=vehicle, group=vehicle)) +
  geom_line() +
  labs(title="Average Monthly Delays by Vehicle", x="Month", y="Monthly Avg. Delay (minutes)") +
  scale_color_manual(values=c("blue","green", "red"), labels=c("Bus","Streetcar", "Subway")) +
  theme_minimal()
```

@@

# Conclusion

# References

@@

# Appendix: TTC Route Nomenclature

>>>>

---
title: "Transit Delays in the Toronto Transit Commission"
subtitle: "An Exploratory Analysis with Open Data Toronto"

author: Luca Carnegie
thanks: "Code and data are available at: https://github.com/lcarnegie/OpenDataToronto. Thank you to Nescafé coffee, Rajan Maghera, Hannah Yu, and Sehar Bajwa for your love and support - I could not have done it without you. Special thanks go to Rohan Alexander for the immense amount of R he taught me in such a short time and his extremely easy-to-navigate textbook (that has a cocktail appendix, fortunately)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
toc: true
fig_caption: yes
number-sections: true
bibliography: references.bib

@@

# Data Analysis {#sec-data}
```{r data import}
#| message: false
#| echo: false
#| label: import data
#| tbl-cap: Sample of Cleaned Delay Data
#| fig-align: center

@@

## Reason of Delay {#subsec1-data}
```{r ggplot Common Delay Reasons}
#| message: false
#| echo: false
#| label: common-reasons-delay
#| tbl-cap: Most Common Reasons for Delay on TTC

@@

```
Subways didn't have any reason for delay, yet have the most incidence of delay. Was dropped from the data.

## Day of Week {#subsec2-data}
```{r ggplot Day of Week with Delay Incidence}
#| message: false
#| echo: false
#| label: transport-delay-time
#| tbl-cap: Most Common Reasons for Delay on TTC
#Day of week with most incidence of delay

@@

num_incidents_vehicle |>
  ggplot(mapping = aes(x = Day, y = num_incidents)) +
  geom_point() +
  geom_line(group = 1) +
  labs(title="Number of Incidents by Day of Week", x="Day of Week", y="Incident Count") +
  theme_minimal() +
  ylim(0, max(num_incidents_vehicle$num_incidents) + 1) +
  theme(plot.title.position = "plot",
        plot.title = element_text(hjust = 0.5))

@@

# Convert 'Day' column to a factor with specific levels
dow_data$Day <- factor(dow_data$Day, levels = days_order, ordered = TRUE)

@@

#plot
avg_time_delayed_vehicle |>
  ggplot(mapping = aes(x = Day, y = `avg_delay`)) +
  geom_point() +
  geom_line(group = 1) + 
  labs(title="Average Delay Time by Day of Week", x="Day of Week", y="Average Delay (mins)") +
  theme_minimal() +
  ylim(0, max(avg_time_delayed_vehicle$avg_delay) + 1) + 
  theme(plot.title.position = "plot",
        plot.title = element_text(hjust = 0.5))
```

avg_time_delayed_vehicle |>
  ggplot(mapping = aes(x = vehicle, y = avg_delay)) +
  geom_col() + 
  labs(title="Average Delay by Vehicle", x="Vehicle Type", y="Average Delay (mins)") +
  theme_minimal() +
  theme(plot.title.position = "plot",
      plot.title = element_text(hjust = 0.5))
```
```{r ggplot transport with most incidence of delays }
#| message: false
#| echo: false
#| label: most-incidents-vehicle

num_incidents_vehicle |>
  ggplot(mapping = aes(x = vehicle, y = num_incidents)) +
  geom_col() + 
  labs(title="Number of Incidents by Vehicle", x="Vehicle Type", y="Number of Incidents") +
  theme_minimal() + 
   theme(plot.title.position = "plot",
        plot.title = element_text(hjust = 0.5))

## Time of Year {#subsec4-data}
```{r ggplot incidence change monthly}
#| message: false
#| echo: false
#| warning: false
#| label: monthly-delay-incidence
#| tbl-cap: Delay Incidence by Month
#Line chart of how avg. delay time changes over the months

#Calculate avg. delays 
delay_incidence <- month_dataset |> group_by(month, vehicle) |> summarise(incidence = n(), .groups = "drop")

# convert month to factor (categorical data)
delay_incidence$month <- factor(delay_incidence$month, levels = 1:12, labels = month.abb[1:12])

# plot the graph
delay_incidence |>
  ggplot(aes(x=month, y= `incidence`, color=vehicle, group=vehicle)) +
  geom_line() +
  labs(title="Delay Count per Month, by Vehicle", x="Month", y="Count of Delays") +
  scale_color_manual(values=c("blue","green", "red"), labels=c("Bus","Streetcar", "Subway")) +
  theme_minimal()
```
```{r ggplot avg delay change monthly}
#| message: false
#| echo: false
#| warning: false
#| label: monthly-delay-times
#| tbl-cap: Average Monthy Delay Times
#Line chart of how avg. delay time changes over the months

@@

# plot the graph
avg_delay |>
  ggplot(aes(x=month, y=`Min Delay`, color=vehicle, group=vehicle)) +
  geom_line() +
  labs(title="Monthly Average Delay Time, by Vehicle", x="Month", y="Monthly Avg. Delay (minutes)") +
  scale_color_manual(values=c("blue","green", "red"), labels=c("Bus","Streetcar", "Subway"))+
  theme_minimal() +
  theme(plot.title.position = "plot",
        plot.title = element_text(hjust = 0.5))
```

## Line/Route {#subsec5-data}
```{r ggplot top 10 routes with longest avg. delay}
#| message: false
#| echo: false
#| label: lines-longest-delays
#| tbl-cap: Incidence of Delay, by Route/Line

```
```{r ggplot top 10 routes with most delay incidence}
#| message: false
#| echo: false
#| label: lines-most-delays
#| tbl-cap: Most Common Reasons for Delay on TTC

@@

# Appendix 
# References

@@
